---
title: "EnrichKit enrichment workflow"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{EnrichKit enrichment workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r setup}
library(EnrichKit)
```

## Build a pathway database

```{r}
gene_sets <- list(
  KEGG_MAPK_SIGNALING_PATHWAY = c("BRAF", "MAP2K1", "MAP2K2", "MAPK1", "MAPK3"),
  BIOCARTA_MTOR_PATHWAY = c("PIK3CA", "AKT1", "MTOR"),
  REACTOME_CELL_CYCLE_CHECKPOINTS = c("CDK1", "CDK2", "RB1")
)

pathway_db <- make_pathway_db(
  gene_sets,
  database = c(
    KEGG_MAPK_SIGNALING_PATHWAY = "KEGG",
    BIOCARTA_MTOR_PATHWAY = "BIOCARTA",
    REACTOME_CELL_CYCLE_CHECKPOINTS = "REACTOME"
  ),
  source = "example_pathway_database",
  min_size = 2
)
```

## Match to an analysis background

The background should be the interrogated feature universe for the analysis.

```{r}
interrogated_genes <- c("BRAF", "MAP2K1", "MAPK1", "MAPK3", "AKT1", "MTOR", "TP53", "CDK1")
matched_db <- match_pathway_background(pathway_db, interrogated_genes, min_size = 2)
pathway_gene_counts(matched_db)
pathway_matching_summary(matched_db)
```

## Fisher enrichment

```{r}
hits <- c("BRAF", "MAP2K1", "MAPK1")

fisher_res <- fisher_enrichment(
  hits = hits,
  pathway_db = matched_db,
  background = interrogated_genes,
  alternative = "greater"
)

fisher_res
```

FDR is database-wise by default. Use `fdr_by_database = FALSE` for a global
adjustment.

## Wilcoxon enrichment

```{r}
scores <- c(
  BRAF = 2.4,
  MAP2K1 = 1.7,
  MAPK1 = 1.3,
  MAPK3 = 1.1,
  AKT1 = -0.3,
  MTOR = -0.8,
  TP53 = 0.2,
  CDK1 = -1.2
)

wilcox_res <- wilcox_enrichment(
  feature_scores = scores,
  pathway_db = matched_db,
  background = interrogated_genes,
  alternative = "greater"
)

wilcox_res
```

## Thresholded Wilcoxon enrichment

Score-thresholded Wilcoxon enrichment first
restricts the scored background to features above an absolute score threshold.

```{r}
thresholded_res <- thresholded_wilcox_enrichment(
  feature_scores = scores,
  pathway_db = matched_db,
  threshold = 1,
  alternative = "greater"
)

thresholded_res
```

## Rank-shift Wilcoxon enrichment

The rank-shift Wilcoxon method from Xiaoyu Song converts scores to ranks and
adds a directional rank margin to the non-pathway background before one-sided
testing.

```{r}
rank_shift_res <- rank_shift_wilcox_enrichment(
  feature_scores = scores,
  pathway_db = matched_db,
  background = interrogated_genes,
  shift_fraction = 0.20,
  direction = "both"
)

rank_shift_res
```

## Redundancy reduction

```{r}
reduced <- reduce_redundant_pathways(
  wilcox_res,
  as_gene_sets(matched_db),
  jaccard_cutoff = 0.8
)

reduced[, c("pathway", "fdr", "redundant", "representative_pathway")]
```

## SUMER handoff

Pathway lists are often redundant. EnrichKit includes cascade-threshold
consolidation directly, and it also prepares files for SUMER, an external
pathway-module consolidation package from the Zhang lab:
https://github.com/bzhanglab/sumer.

```{r}
cascade <- cascade_threshold_consolidation(
  wilcox_res,
  as_gene_sets(matched_db),
  min_new_members = 2,
  row_order = "input"
)

cascade$results[, c("pathway", "cascade_n_new_members", "cascade_retained")]
```

For SUMER, EnrichKit prepares the input files, writes an editable config
template, can call SUMER when `run = TRUE`, and reads module output after SUMER
has run.

```{r eval = FALSE}
sumer_job <- sumer_workflow(
  enrichment = wilcox_res,
  gene_sets = as_gene_sets(matched_db),
  out_prefix = "pathway_enrichment_sumer",
  weight_col = "signed.fdr",
  top_num = 100,
  platform_name = "pathway_enrichment",
  platform_abbr = "pathway",
  run = FALSE
)

sumer_job$prep$gmt_file
sumer_job$prep$data_file
sumer_job$config_file
```

The files are:

- `*_pathways.gmt`: pathway definitions.
- `*_data.txt`: pathway weights.
- `*_sumer_config.json`: editable SUMER config template.

After editing the config for the local SUMER installation:

```{r eval = FALSE}
sumer_job <- sumer_workflow(
  enrichment = wilcox_res,
  gene_sets = as_gene_sets(matched_db),
  out_prefix = "pathway_enrichment_sumer",
  weight_col = "signed.fdr",
  config_file = "pathway_enrichment_sumer_sumer_config.json",
  output_name = "pathway_enrichment_sumer_output",
  run = TRUE
)

sumer_job$modules$module_summary
```

## Supplementary table

```{r}
supp <- format_enrichment_supplement(wilcox_res, analysis_name = "MAPK pathway example")
supp
```
